Linear Ordered Attribute Grammar scheduling using SAT-solving

### 2014-10-12 paper/paper.tex
Een aantal plaatjes met begeleidende tekst en commentaar.

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\input{./formats.tex}

\title{Linear Ordered Attribute Grammar scheduling using SAT-solving}
\author{Jeroen Bransen\inst{1}  \and L. Thomas van Binsbergen\inst{2,1} \and Koen Claessen\inst{3} \and Atze Dijkstra\inst{1}}
\institute{Utrecht University, Utrecht, The Netherlands, \email{\{J.Bransen,atze\}@uu.nl}
\and Royal Holloway, University of London, London, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
\and Chalmers University of Technology, Gothenburg, Sweden, \email{koen@chalmers.se}}

@@

In this paper, we show how to encode linear ordered attribute grammar
scheduling as a SAT-problem. Cycle-freeness of the dependency graphs
is expressed by making the graphs chordal, which is a new approach for
encoding such graph property in SAT.

There are two main advantages to using a SAT-solver for scheduling:
(1) the scheduling algorithm runs much faster than existing scheduling
algorithms on real-world examples, and (2) by adding extra constraints
we obtain fine-grained control over the resulting schedule, thereby
enabling new scheduling optimizations.

@@

\section{Linear Ordered Attribute Grammars}
\input{./loags.tex}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usetikzlibrary{arrows, automata}
\input{./formats.tex}

\title{Linear Ordered Attribute Grammar scheduling using SAT-solving}
\author{Jeroen Bransen\inst{1}  \and L. Thomas van Binsbergen\inst{2,1} \and Koen Claessen\inst{3} \and Atze Dijkstra\inst{1}}
\institute{Utrecht University, Utrecht, The Netherlands, \email{\{J.Bransen,atze\}@uu.nl}
\and Royal Holloway, University of London, Egham, UK, \email{Thomas.VanBinsbergen.2014@live.rhul.ac.uk}
\and Chalmers University of Technology, Gothenburg, Sweden, \email{koen@chalmers.se}}

\newif\iffinal\finalfalse
\newcommand{\REM}[3]{\iffinal\else\textcolor{#2}{[#1: #3]}\fi}
\newcommand{\Atze}[1]{\REM{Atze}{red}{#1}}
\newcommand{\Koen}[1]{\REM{Koen}{red}{#1}}
\newcommand{\Jeroen}[1]{\REM{Jeroen}{red}{#1}}
\newcommand{\Thomas}[1]{\REM{Thomas}{red}{#1}}

@@

In this paper, we show how to encode linear ordered attribute grammar
scheduling as a SAT-problem. Cycle-freeness of the dependency graphs
is expressed by making the graphs chordal, which is a new approach for
encoding such graph properties in SAT.

There are two main advantages to using a SAT-solver for scheduling:
(1) the scheduling algorithm runs much faster than existing scheduling
algorithms on real-world examples\Thomas{Waarmee vergelijken we? 
Kastens algoritme is niet voor LOAGs, het voordeel dat onze aanpak geen 
augmenting dependencies nodig heeft zou dan in de vergelijking genoemd moeten 
worden}, and (2) by adding extra constraints
we obtain fine-grained control over the resulting schedule, thereby
enabling new scheduling optimizations.

@@

\section{Attribute Grammar}
An attribute grammars consists of three parts: a context-free grammar 
describing abstract syntax, a set of inherited and synthesized attributes
assigned to every non-terminal of the grammar and a set of semantic function
definitions. A production rule of a context-free grammar has one non-terminal
as its left-hand side and any number of terminals and non-terminals as its
right-hand side. The non-terminals occurring in production rules are considered
to be non-terminal \emph{occurrences}. A non-terminal occurrence of 
non-terminal $X$ automatically obtains the attributes of $X$ as attribute 
occurrences. The semantic function definitions describe how some attribute 
occurrences of a production rule are calculated in terms of other attribute
occurrences and terminal symbols of that rule. For our purposes we can safely
ignore any terminal symbols. Figure \ref{fig:productions} shows
examples of production rules displayed as diagrams.

\mypic{productions}{Diagrams showing productions, $P1: (lhs:X) \rightarrow 
\alpha$, $P2: (lhs:Y) \rightarrow \alpha (x:X) \beta$, $P3: (lhs:Z)
\rightarrow \alpha (y:Y) \beta (x:X) \gamma$, where $\alpha$, $\beta$,
$\gamma$ are arbitrary terminal symbols.}

All non-terminal occurrences are named, e.g. $(x:X)$ is an occurrence of 
non-terminal $X$ named $x$. Synthesized attributes are shown to the right
and inherited attributes are shown to the left of their respective
non-terminal occurrences. The gray attribute occurrences (inherited in the
left-hand side and synthesized in the right-hand side) are input occurrences,
the others are output occurrences.

The input occurrences of one production are the output occurrences of another.
Production rules can be `pasted' together to form a tree\cite{knuth68}. 
This task is performed by a parser forming a derivation 
tree\Thomas{Ik stop hier maar even met typen om 
niet te veel uit te wijden, maar eigenlijk zou ik hier nog willen vertellen
dat: 1) Input occurrences van buitenaf worden aangeleverd, 2) dat om dit
te kunnen doen er wel definities voor output occurrences nodig zijn, 
3) dat de pijlen in het figuur (omgekeerde) dependencies weergeven.}.

\section{Linear Ordered Attribute Grammars}
A linear ordered attribute grammar is an attribute grammar for which we can 
find a evaluation order statically\Thomas{Hier zouden we misschien een 
voorbeeld van een parse tree willen laten zien met een interessante 
evaluation order, helaas zijn dat soort plaatjes veel te groot}. 
That is, based on only the information of
non-terminals and productions from the grammar, we can ensure that an 
evaluation order exists for any valid sentence of the grammar.
We show a static evaluation order exists by constructing evaluation orders 
for all production rules such that they satisfy certain properties.
Most importantly: when pasting together production rules to form a 
derivation tree, the evaluation orders for those trees can not contradict 
each other. Figure \ref{fig:contradiction} shows what can go wrong.

\mypic{contradiction}{The evaluation orders depicted by the arrows
of productions $P4$, $P5$ and $P6$ contradict each other: they can
not be combined without forming a cycle.}

Any non-terminal occurrence of $X$ can be replaced by any
of the productions of $X$. Therefore, when constructing an evaluation
order for a production in which $X$ occurs in the right-hand side,
all the evaluation orders constructed for productions in which 
$X$ occurs in the left-hand side need to be taken account.
Changing the evaluation order of one production rule is likely to demand 
a change in the evaluation order of some other production. This is 
what makes LOAG scheduling NP-hard.

In a production graph we can represent the evaluation order of other graphs
using induced dependencies (graph $ID_P$ from Kastens) by adding an 
arrow $a\rightarrow b$ to a non-terminal $X$ if $X$ occurs in some other
production in which there is a path between its $a$ and $b$.
Figure \ref{fig:idp} shows the induced dependency graph of production
$P4$.

\mypic{idp}{The gray edges are induced dependency, showing that they
evaluation orders of Figure \ref{fig:contradiction} are not compatible.}

The induced dependencies will be the same for every occurrence of every
non-terminal and are gathered in induced dependency graphs for 
non-terminals (Kastens $ID_S$).

We can now look at the problem from two directions: 1) we change the
evaluation order of a production and update graph $ID_S$ which might
then affect other production rules and update until all evaluation
orders are compatible\Thomas{AOAG} or 2) we make the decision at non-terminal 
level and decide an order for evaluating the attributes of a non-terminal 
such that an order at production level is still 
possible\Thomas{LOAG, De intuitie achter dit idee zijn de visit-sequences
en intra-visit dependencies, maar die willen we niet noemen}.

\section{Chordal Graphs}
The chords of a chordal graph can be seen as representing the assignment
to the variables `behind' that chord. If there is a chord $a\leftrightarrow b$
then an assignment $a\leftrightarrow^\rightarrow b$ implies that there is 
no path $b\longrightarrow a$ assuming cycle-freeness. This intuition allows
us to see why all assignments that lead to no cycles of length 3 lead to 
no cycles at all in a chordal graph: every cycle of length $> 3$ is 
split by a chord, so one side of the cycles assignments is represented by
the chord in the other.

\mypic{simplechord}{There can not exists a cycle w-n-e-s without either
a cycle n-s-w or n-e-s.}

\Thomas{In mijn scriptie gebruik ik problem graphs en assignment graphs
om de stap van undirected naar directed graphs te maken. Later kwam ik
echter het begrip `orientation' tegen, met hetzelfde doel.}

\section{Definitions}
\input{./loags.tex}
